title: "Generate cell type frequency barplots v2"
author: "Lukas Simon"

Figure 3a-d
---

Load R packages
```{r}
library(reticulate)
library(ggplot2)
library(gridExtra)
library(reshape2)
```

Load scanpy object in python and extract obs table
```{}
repl_python()
import scanpy as sc
adata = sc.read('integrated_human_dataset.h5ad')
obs = adata.obs
exit
```

Calculate cell type frequencies
```{r}
metadata <- py$obs
asplit <- split(1:nrow(metadata), metadata$patient_id)
celltypes <- unique(metadata$cell_type)
tmp <- do.call(cbind, lapply(asplit, function(x) table(metadata$cell_type[x])[celltypes]))
rownames(tmp) <- celltypes
tmp[which(is.na(tmp))] <- 0
tmp <- t(t(tmp) / colSums(tmp))
freqs <- tmp
study <- rep('Reyfman', ncol(freqs))
study[grep('GSM', colnames(freqs))] <- 'Banovich'
study[grep('muc', colnames(freqs))] <- 'Schiller'
treat <- metadata$health_status[match(colnames(freqs), metadata$patient_id)]
age <- metadata$Age[match(colnames(freqs), metadata$patient_id)]
gender <- metadata$Sex[match(colnames(freqs), metadata$patient_id)]
```

Generate boxplots
```{r}
tmp <- data.frame(t(freqs), study, treat)
aframe <- melt(tmp, measure.vars = setdiff(colnames(tmp), c('study', 'treat')))
old_labels <- unique(paste(aframe$treat, aframe$variable, sep = '.'))
new_labels <- do.call(c, lapply(as.character(unique(aframe$variable)), function(x) c(x, '')))
p_schiller <- ggplot(aframe[which(aframe$study == 'Schiller'),], aes(x = interaction(treat, variable), value, color = treat)) +
    geom_boxplot(outlier.shape = NA) + geom_point(alpha = 0.5, size = 1) +
    theme_bw() + coord_flip() +
    scale_x_discrete(breaks = old_labels, labels = new_labels) +
    ylab('Frequency') + xlab('Cell type') + ggtitle('Schiller') +
  theme(legend.position = "none")
  
p_banovich <- ggplot(aframe[which(aframe$study == 'Banovich'),], aes(x = interaction(treat, variable), value, color = treat)) +
    geom_boxplot(outlier.shape = NA) + geom_point(alpha = 0.5, size = 1) +
    theme_bw() + coord_flip() +
    scale_x_discrete(breaks = old_labels, labels = new_labels) +
    ylab('Frequency') + xlab('Cell type') + ggtitle('Banovich') +
  theme(legend.position = "none")
p_reyfman <- ggplot(aframe[which(aframe$study == 'Reyfman'),], aes(x = interaction(treat, variable), value, color = treat)) +
    geom_boxplot(outlier.shape = NA) + geom_point(alpha = 0.5, size = 1) +
    theme_bw() + coord_flip() +
    scale_x_discrete(breaks = old_labels, labels = new_labels) +
    ylab('Frequency') + xlab('Cell type') + ggtitle('Reyfman') +
  theme(legend.position = "none")
p_final <- grid.arrange(p_schiller, p_banovich, p_reyfman, ncol = 3)
ggsave(p_final, file = 'Cell_type_frequencies.pdf', width = 16, height = 8)
```

Calculate foldchanges
```{r}
fcs <- unlist(lapply(split(1:nrow(aframe), aframe$variable), function(x){
  subm <- aframe[x,]
  ctrl <- mean(subm$value[which(subm$treat == 'control donor')])
  ild <- mean(subm$value[which(subm$treat == 'endstage lung fibrosis')])
  (ild + 0.001)/(ctrl + 0.001)
}))
```

Foldchanges for cell types used in paper
```{r}
fcs[c('Aberrant.basaloid.cells', 'mDC1', 'Basal.cells', 'Club.cells',
      'AM.activated', 'AT.1.cells', 'B.cells', 'AT.2.cells', 'pDC', 'Myofibroblasts')]
```

Perform PCA
```{r}
asplit <- split(1:ncol(freqs), study)
aframe <- do.call(rbind, lapply(asplit, function(x){
  prop <- sqrt(freqs[,x])
  pca <- prcomp(t(prop))
  umap::umap(pca$x[,1:5])
  udata <- umap::umap(pca$x[,1:5])
  data.frame(pca$x[,1:5],
             udata$layout,
             treat = treat[x], study = study[x],
             age = age[x], gender = gender[x])
}))
ggplot(aframe, aes(PC1, PC2, color = treat)) +
  facet_wrap(~ study) +
  geom_point() + theme_bw()
```

Plot PCA based on cell type frequencies
```{r}
prop <- freqs
prop <- prop[which(apply(prop, 1, var) > 0),] 
pca <- prcomp(t(prop), scale. = T)
aframe <- data.frame(pca$x[,1:5],
                     treat, study, age, gender)
ggplot(aframe, aes(PC1, PC2, color = treat)) +
  stat_ellipse() + geom_point(data = aframe, aes(shape = study)) +
  theme_bw() + ggtitle('PCA of cell type frequencies')
ggsave('Cell_type_freq_PCA.pdf', width = 8, height = 6)
ggplot(aframe, aes(treat, PC2, fill = treat)) +
  facet_wrap(~ study) +
  geom_boxplot() + geom_point() + theme_bw()
ggsave('Cell_type_freq_PC2.pdf', width = 6, height = 6)
```

Run RF analysis (5 fold cross-validation)
```{r}
library(randomForest)
prop <- sqrt(freqs)
aframe <- data.frame(treat, t(prop))
k <- 5
asplit <- split(colnames(prop), 1:k) 
orig <- unlist(lapply(asplit, function(x){
  test <- x
  training <- setdiff(colnames(prop), test)
  #model <- lda(treat ~ ., data = aframe[training,])
  #predictions <- predict(model, data.frame(aframe[test,]))
  #mean(predictions$class == aframe[test, 'treat'])
  model <- randomForest(treat~., data = aframe[training,])
  predictions <- predict(model, data.frame(aframe[test,]))
  mean(predictions == aframe[test, 'treat'])
}))
aframe <- data.frame(treat = sample(aframe$treat), t(prop))
random <- unlist(lapply(asplit, function(x){
  test <- x
  training <- setdiff(colnames(prop), test)
  #model <- lda(treat ~ ., data = aframe[training,])
  #predictions <- predict(model, data.frame(aframe[test,]))
  #mean(predictions$class==aframe[test, 'treat'])
  
  model <- randomForest(treat~., data = aframe[training,])
  predictions <- predict(model, data.frame(aframe[test,]))
  mean(predictions == aframe[test, 'treat'])
}))
aframe <- data.frame(accuracies = c(orig, random), type = c(rep('orig', k), rep('random', k)))
ggplot(aframe, aes(type, accuracies)) + geom_boxplot() + geom_point() + theme_bw() +
  ggtitle('5 fold cross-validation') + ylab('Accuracy') + xlab('')
ggsave('RF_cell_type_freq_accuracies.pdf', height = 6, width = 6)
aframe <- data.frame(treat, t(prop))
model <- randomForest(treat~., data = aframe)
imp <- as.numeric(model$importance)
names(imp) <- rownames(model$importance)
pdf('RF_cell_type_freq_imporance.pdf', height = 8, width = 6)
par(mar = c(10, 5, 5, 5))
barplot(tail(sort(imp), 10), las = 2, ylab = 'RF imporance')
dev.off()
```

Run LDA analysis (leave one study out)
```{r}
prop <- sqrt(freqs)
prop <- prop[which(apply(prop, 1, var) > 0),] 
aframe <- data.frame(treat, t(prop))
asplit <- split(colnames(prop), study) 
orig <- unlist(lapply(asplit, function(x){
  test <- x
  training <- setdiff(colnames(prop), test)
  #model <- lda(treat~., data = aframe[training,])
  model <- randomForest(treat~., data = aframe[training,])
  predictions <- predict(model, data.frame(aframe[test,]))
  mean(predictions == aframe[test, 'treat'])
}))
aframe <- data.frame(accuracies = c(orig, random), type = c(rep('orig', k), rep('random', k)))
ggplot(aframe, aes(type, accuracies)) + geom_boxplot() + geom_point() + theme_bw() +
  ggtitle('5 fold cross-validation') + ylab('Accuracy') + xlab('')
```

Plot variances
```{r}
asplit <- split(1:ncol(freqs), paste(study, treat, sep = '|'))
tmp <- do.call(rbind, lapply(names(asplit), function(x){
  vars <- apply(freqs[, asplit[[x]]], 1, var)
  means <- apply(freqs[, asplit[[x]]], 1, mean)
  data.frame(treat = strsplit(x, '|', fixed = T)[[1]][2],
             study = strsplit(x, '|', fixed = T)[[1]][1],
             vars,
             means,
             x = names(vars))
}))
aframe <- tmp
aframe$coef_var <- aframe$vars/aframe$means
ggplot(aframe, aes(x=treat, y=coef_var, fill=treat)) +
  facet_wrap(~ study) +
  geom_boxplot()
aframe$coef_var[which(aframe$treat == 'control donor')] <- aframe$coef_var[which(aframe$treat == 'control donor')] * (-1)
ggplot(aframe, aes(x=x, y=coef_var, fill=treat)) +
  facet_wrap(~ study) +
  geom_bar(stat = 'identity') +
  coord_flip() + theme_bw()
```


Plot variances (balanced groups)
```{r}
asplit <- split(1:ncol(freqs), study)
tmp <- do.call(rbind, lapply(names(asplit), function(nom){
  x <- asplit[[nom]]
  study <- nom
  asplit2 <- split(colnames(freqs)[x], treat[x])
  n <- min(unlist(lapply(asplit2, length)))
  do.call(rbind, lapply(names(asplit2), function(nom2){
    y <- asplit2[[nom2]]
    treat <- nom2
    ok <- sample(y, n)
    vars <- apply(freqs[, ok], 1, var)
    means <- apply(freqs[, ok], 1, mean)
    data.frame(treat,
               study,
               vars,
               means,
               x = 1:length(vars))  
  }))
}))
aframe <- tmp
aframe$coef_var <- aframe$vars/aframe$means
ggplot(aframe, aes(x=treat, y=coef_var, fill=treat)) +
  facet_wrap(~ study) +
  geom_boxplot()
```

```{python}
am = adata[adata.obs['cell_type'] == 'AM']
patients = list(am.obs['patient_id'].unique())
cells = [am[am.obs['patient_id'] == patient].obs_names for patient in patients]
cols = []
for patient in list(patients):
    for x in np.arange(0, 3):
        cols.append(patient + '_' + str(x))
        
means.columns = cols
```

```{r}
means <- py$means
means <- means[which(rowSums(means) > 0),]
means <- means[which(apply(means, 1, function(x) sum(x > 0)) >= 3),]
patient <- colnames(means)
patient <- unlist(lapply(patient, function(x) strsplit(x, '_', fixed = T)[[1]][1]))
pvals <- t(apply(means, 1, function(x){
  coefficients(summary(lm(x ~ patient)))[-1, 4]
}))
unique_genes <- lapply(1:ncol(pvals), function(x){
 ok1 <- which(apply(pvals[, -x], 1, function(y) sum(y > 0.05)) == ncol(pvals) - 1)
 ok1 <- rownames(pvals)[ok1]
 names(head(sort(pvals[ok1, x]), 10))
})
tmp <- means[unlist(unique_genes), ]
pheatmap(tmp, scale = 'row', cluster_rows = F, cluster_cols = F, breaks = seq(-2, 2, length = 101))